Appendix S1

### 2020-06-10 analysis/paper/appendix_s1.Rmd
revised the language explaning our selection process. Dropped land classification metrics from MIC table. Added colophon to the bottom.

A key step in developing a QRF model to predict fish capacities was selecting the habitat covariates to include in the model. Random forest models naturally incorporate interactions between correlated covariates, which is essential since nearly all habitat variables are considered correlated to one degree or another, however, we aimed to avoid overly redundant variables (i.e., variables that measure similar aspects of the habitat). Further, including too many covariates can result in overfitting of the model (e.g., including as many covariates as data points).

To prevent overfitting, we pared down the more than 100 metrics generated by the CHaMP protocol describing the quantity and quality of fish habitat for each survey site. To assist in determining the habitat metrics to include in the QRF model, we used the Maximal Information-Based Nonparametric Exploration (MINE) class of statistics [@Reshef2011] to determine those habitat characteristics (covariates) most highly associated with observed parr densities. We calculated the maximal information coefficient (MIC), using the R package *minerva* [@R-minerva], to measure the strength of the linear or non-linear association between two variables [@Reshef2011]. The MIC value between each of the measured habitat characteristics and parr density was used to inform decisions on which habitat covariates to include in the QRF parr capacity model.

Habitat metrics were first grouped into broad categories that included channel unit, complexity, cover, disturbance, riparian, size, substrate, temperature, water quality, and woody debris. Habitat metrics measuring volume or area were scaled to the wetted area of each site. Within each category, metrics were ranked according to their MIC value (Table \@ref(tab:mic-tab) and Figure \@ref(fig:mine-facet-fig)). Our strategy was to select one or two variables with the highest MIC score within each category so that covariates describe different aspects of rearing habitat (e.g., substrate, temperature, etc.). Additionally, we attempted to avoid covariates that were highly correlated (**Figure ??**) while including covariates that can be directly influenced by restoration actions or have been shown to impact salmonid juvenile density.

<!-- talk about how we chose not to use categorical variables like channle type or valley class, because we expected other metrics to capture the difference between those categories -->

@@

```{r mic-tab}
mine_tab_df = mine_res %>%
  group_by(MetricCategory) %>%
  slice(1:10) %>%
  select(Category = MetricCategory,
         Name,
         Abbrv = Metric,
         MIC,
         selected)

@@

```{r corr-fig, eval = F, results='hide', fig.width = 5, fig.height = 5, fig.show = "hold", fig.cap = "Correlation plots of metrics, facted by habitat category."}
hab_corr$catg_cor[[1]][!sapply(hab_corr$catg_cor[[1]], is.null)] %>%
  # map(.f = function(x) {
  #   rownames(x) <- tibble(Metric = rownames(x)) %>%
  #     left_join(hab_dict %>%
  #                 select(Metric = ShortName,
  #                        Name) %>%
  #                 distinct()) %>%
  #     pull(Name)
  #   
  #   # colnames(x) <- tibble(Metric = colnames(x)) %>%
  #   #   left_join(hab_dict %>%
  #   #               select(Metric = ShortName,
  #   #                      Name) %>%
  #   #               distinct()) %>%
  #   #   pull(Name)
  #   return(x)
  # }) %>%
    map(.f = corrplot::corrplot.mixed,
        upper = 'ellipse',
        tl.pos = 'lt',
        order = 'FPC')
```

```{r corr-fig2, results='hide', fig.width = 5, fig.height = 5, fig.show = "hold", fig.cap = "Correlation plots of metrics, facted by habitat category."}
cor_plot_df = tibble(Category = names(hab_corr$catg_cor[[1]][!sapply(hab_corr$catg_cor[[1]], is.null)]),
                     cor_mat = hab_corr$catg_cor[[1]][!sapply(hab_corr$catg_cor[[1]], is.null)]) %>%
  mutate(cor_plot = map2(cor_mat,
                         Category,
                         .f = function(x, y) {
                           ggcorrplot::ggcorrplot(x,
                                                  method = 'circle',
                                                  hc.order = T,
                                                  type = 'lower',
                                                  title = y,
                                                  colors = c('blue4', 'white', 'red3'))
                         }))

@@

>>>>

A key step in developing our QRF model to predict fish capacities was selecting the habitat covariates to include in the model. Random forest models naturally incorporate interactions between correlated covariates, which is essential since nearly all habitat variables are considered correlated to one degree or another, however, we aimed to avoid overly redundant variables (i.e., variables that measure similar aspects of the habitat). Further, including too many covariates can result in overfitting of the model (e.g., including as many covariates as data points). Our goal was to select a group of covariates that captured as many different aspects of the stream habitat (e.g. substrate, flow, riparian condition, channel unit configuration, etc.) as possible, while still holding information about fish densities.

To prevent overfitting, we pared down the more than 100 metrics generated by the CHaMP protocol describing the quantity and quality of fish habitat for each survey site. Habitat metrics were first grouped into broad categories that included channel unit, complexity, cover, riparian areas, side channels, size, substrate, temperature, water quality, and woody debris. Habitat metrics measuring any large wood volume were scaled by the site length (in 100 m units). To assist in determining the habitat metrics to include in the QRF model, we used the Maximal Information-Based Nonparametric Exploration (MINE) class of statistics [@Reshef2011] to determine those habitat characteristics (covariates) most highly associated with the log of observed parr densities. We calculated the maximal information coefficient (MIC), using the R package *minerva* [@R-minerva], to measure the strength of the linear or non-linear association between two variables [@Reshef2011]. MIC is a measure of correlation that incorporates potential non-linear associations, so for example if there is an expontential association the MIC value could be high, even when the standard correlation coefficient is low. We excluded categorical variables such as channel type (e.g. meandering, pool-riffle, plane-bed, etc.) because we assumed that other quantitative metrics would capture the differences between those qualitative categorical metrics.

Within each category, metrics were ranked according to their MIC value (Table \@ref(tab:mic-tab) and Figure \@ref(fig:mine-facet-fig)). The MIC value between each of the measured habitat characteristics and parr density was used to inform decisions on which habitat covariates to include in the QRF parr capacity model. We selected one or two variables amongst those with the highest MIC scores within each category, attempting to avoid covariates that were too highly correlated (Figure \@ref(fig:corr-fig)), while focusing on covariates we thought could influence fish behavior. For example, cumulative drainage area, mean annual flow and observed discharge are all highly correlated, but fish really only experience the observed flow, so we chose to include that metric in our QRF model. We also tried to include covariates that can be directly influenced by restoration actions or have been shown to impact salmonid juvenile density. Finally, we attempted to avoid metrics with too many missing values, or too many all zero values, in the data set, as well as metrics that may have too much observer error [@Rosgen2018].

@@

```{r mic-tab}
mine_tab_df = mine_res %>%
  filter(MetricCategory != 'Land Classification') %>%
  group_by(MetricCategory) %>%
  slice(1:10) %>%
  select(Category = MetricCategory,
         Name,
         Abbrv = Metric,
         MIC,
         perc_non0,
         selected)

@@

```{r corr-fig1, eval = F, results='hide', fig.width = 5, fig.height = 5, fig.show = "hold", fig.cap = "Correlation plots of metrics, facted by habitat category."}
hab_corr$catg_cor[[1]][!sapply(hab_corr$catg_cor[[1]], is.null)] %>%
  # map(.f = function(x) {
  #   rownames(x) <- tibble(Metric = rownames(x)) %>%
  #     left_join(hab_dict %>%
  #                 select(Metric = ShortName,
  #                        Name) %>%
  #                 distinct()) %>%
  #     pull(Name)
  #   
  #   # colnames(x) <- tibble(Metric = colnames(x)) %>%
  #   #   left_join(hab_dict %>%
  #   #               select(Metric = ShortName,
  #   #                      Name) %>%
  #   #               distinct()) %>%
  #   #   pull(Name)
  #   return(x)
  # }) %>%
    map(.f = corrplot::corrplot.mixed,
        upper = 'ellipse',
        tl.pos = 'lt',
        order = 'FPC')
```

```{r corr-fig, results='hide', fig.width = 5, fig.height = 5, fig.show = "hold", fig.cap = "Correlation plots of metrics, facted by habitat category."}
cor_plot_df = tibble(Category = names(hab_corr$catg_cor[[1]][!sapply(hab_corr$catg_cor[[1]], is.null)]),
                     cor_mat = hab_corr$catg_cor[[1]][!sapply(hab_corr$catg_cor[[1]], is.null)]) %>%
  mutate(cor_plot = map2(cor_mat,
                         Category,
                         .f = function(x, y) {
                           ggcorrplot::ggcorrplot(x,
                                                  method = 'circle',
                                                  hc.order = T,
                                                  type = 'lower',
                                                  title = y,
                                                  colors = c('blue4', 'white', 'red3'))
                         }))

@@

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```
